`r glue::glue('Initial inferCNV results for {params$library_id}')`

### 2025-04-14 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
progress

# note these tsvs already have an indicator column for the reference
# the `subcluster` contains the inferCNV-calculated clusters
infercnv_df <- readr::read_tsv(immune_hmm_tsv) |>
  dplyr::bind_rows(
    readr::read_tsv(subset_hmm_tsv)
  ) |>
  dplyr::mutate(
    barcodes = stringr::str_remove(cell_id, glue::glue("{params$library_id}-")),
    # add indicator columns for whether the cell is reference or query
    is_immune_reference = barcodes %in% immune_ref_barcodes,
    is_subset_reference = barcodes %in% subset_ref_barcodes
  )

# Read SCE and create data frame with full results
cnv_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
    \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
  ) |>
  # add in hmm results
  dplyr::left_join(infercnv_df, by = "barcodes") |>
  # add in cell type results
  dplyr::left_join(celltype_df) |>
  # remove MT and GL chromosomes
  dplyr::select(
    -contains("chrMT"),
    -contains("chrGL")
  )
```

@@

![InferCNV heatmap with immune-subset reference](`r immune_png`)

@@

```{r}
# grab any columns that have has_cnv
has_cnv_cols <- names(cnv_df)[stringr::str_starts(names(cnv_df), "has_cnv")]

per_cell_cnv_df <- cnv_df |>
  dplyr::mutate(total_cnv_per_cell = rowSums(dplyr::across(dplyr::all_of(has_cnv_cols)))) |>
  dplyr::rename(infercnv_cluster = subcluster)
```

```{r fig.width = 10, fig.height = 5}
all_plot <- per_cell_cnv_df |>
  dplyr::filter(normal_reference == "all-immune") |>
  dplyr::rename(in_reference = is_immune_reference) |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = in_reference) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-all reference"
  ) +
  theme(legend.position = "bottom")

subset_plot <- per_cell_cnv_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  dplyr::rename(in_reference = is_subset_reference) |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = in_reference) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-subset reference"
  ) +
  theme(legend.position = "bottom")

wrap_plots(all_plot, subset_plot, guides = "collect") +
  plot_annotation(title = "Total CNVs per cell") &
  theme(legend.position = "bottom")
```

@@

Next, we'll look at the per-cell distributions of CNVs across cell types.
Specifically, we'll look at both the cell types from the `cell-type-consensus` module and the `cell-type-ewings` module separately.
Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
consensus_violin <- ggplot(per_cell_cnv_df) +
  aes(
    x = consensus_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-consensus") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ewings_violin <- ggplot(per_cell_cnv_df) +
  aes(
    x = ewing_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-ewings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
per_cell_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(color = "annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```
Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.

```{r fig.width=10}
ggplot(per_cell_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell") +
  umap_theme
```

@@

Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV. 
To see whether we'd like to further tune the clustering parameters, we'll get a size of some of their properties here:

How many cells are there per cluster?
Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.

```{r}
cells_per_cluster <- per_cell_cnv_df |>
  dplyr::count(infercnv_cluster, normal_reference)

ggplot(cells_per_cluster) +
  aes(
    x = infercnv_cluster, # keep in default order so "reference" and "unknown" are grouped
    y = n
  ) +
  geom_col() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

We want to look across chrs now, specifically:

@@

```{r}
cnv_validation_df <- per_cell_cnv_df |>
  dplyr::select(
    barcodes,
    infercnv_cluster,
    total_cnv_per_cell,
    normal_reference,
    is_immune_reference,
    is_subset_reference,
    all_of(
      c(
        glue::glue("proportion_scaled_cnv_chr{1:22}"),
        glue::glue("proportion_scaled_loss_chr{1:22}"),
        glue::glue("proportion_scaled_dupli_chr{1:22}")
      )
    )
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
```

>>>>

# note these tsvs already have an indicator column for the reference
# the `subcluster` contains the inferCNV-calculated clusters
infercnv_df <- readr::read_tsv(immune_hmm_tsv) |>
  dplyr::bind_rows(
    readr::read_tsv(subset_hmm_tsv)
  ) |>
  tidyr::separate(cell_id, into = c("library_id", "barcodes"))

# Read SCE and save UMAP coordinates
umap_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  dplyr::select(
    barcodes,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )
```

@@

![InferCNV heatmap with immune-subset reference](`r subset_png`)

@@

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    normal_reference,
    subcluster,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each cell, but separately for each run of inferCNV (aka each normal reference)
  dplyr::group_by(barcodes, cell_id, normal_reference) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  # add indicator columns for whether the cell is reference or query
  dplyr::mutate(
    immune_reference_cell = barcodes %in% immune_ref_barcodes,
    subset_reference_cell = barcodes %in% subset_ref_barcodes
  ) |>
  # join in other data
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df)
```

```{r fig.width = 10, fig.height = 5}
all_plot <- has_cnv_df |>
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = immune_reference_cell) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-all reference"
  ) +
  theme(legend.position = "bottom")

subset_plot <- has_cnv_df |>
  dplyr::filter(normal_reference == "subset-immune") |>
  ggplot() +
  aes(x = total_cnv_per_cell, fill = subset_reference_cell) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Immune-subset reference"
  ) +
  theme(legend.position = "bottom")

wrap_plots(all_plot, subset_plot) +
  plot_annotation(title = "Total CNVs per cell")
```

@@

Next, we'll look at the per-cell distributions of CNVs across cell types.
Specifically, we'll look at both the cell types from the `cell-type-consensus` module and the `cell-type-ewings` module separately.

Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
In addition, comparing distributions between `all-immune` and `subset-immune` may be helpful to identify the optimal normal reference strategy.

```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# drop NAs from reference cells from other libraries
query_has_cnv_df <- has_cnv_df |>
  tidyr::drop_na(ewing_annotation, consensus_annotation)

consensus_violin <- ggplot(query_has_cnv_df) +
  aes(
    x = consensus_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-consensus") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ewings_violin <- ggplot(query_has_cnv_df) +
  aes(
    x = ewing_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-ewings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
query_has_cnv_df |>
  # arbitrary filtering, this is just data
  dplyr::filter(normal_reference == "all-immune") |>
  ggplot() +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  facet_wrap(vars(normal_reference)) +
  labs(color = "annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```
Next, the total CNVs per cell; ideally, higher values will overlap with regions in the UMAP containing putative tumor cells.

```{r fig.width=10, warning = FALSE}
# add clusters into data frame
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    normal_reference,
    infercnv_cluster = subcluster
  ) |>
  dplyr::inner_join(
    has_cnv_df,
    by = c("barcodes", "normal_reference")
  )

# noting we expect warnings with this plot since there's no UMAP coordinates
# for cells from other libraries in the reference
ggplot(has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = total_cnv_per_cell) +
  geom_point(size = 0.1, alpha = 0.3) +
  scale_color_viridis_c() +
  facet_wrap(vars(normal_reference)) +
  labs(title = "Total CNV per cell") +
  umap_theme
```

@@

Part of `inferCNV`'s algorithm involves clustering the data, and poor clusters have the potential to lead to inaccurate inferences of CNV. 
To see whether we'd like to further tune the clustering parameters, we'll get a sense of their size and rough composition here.

First, how many cells are there per cluster?
Clusters labeled `reference_` contain only normal reference cells, and clusters labeled `unknown_` contain the query cells.

```{r}
ggplot(has_cnv_df) +
  # keep in default order so "reference" and "unknown" are grouped
  aes(x = infercnv_cluster) +
  geom_bar() +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Let's also look at the distributions of cell types in clusters.
Is there a trend for which cell types tend to show up in which cluster?
We'll plot this for both the consensus and ewing annotations, considering only the top 7 cell types for visual ease.

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = infercnv_cluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Consensus cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = infercnv_cluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Ewing cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Let's also look more specifically at the `reference` clusters, where we might expect some batch effects since cells were taken from several different libraries.
Do originating libraries cluster together?

```{r fig.width=10}
all_reference_clusters <- has_cnv_df |>
  # bring in library_id
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(
    normal_reference == "all-immune",
    immune_reference_cell
  )
all_plot <- ggplot(all_reference_clusters) +
  aes(x = infercnv_cluster, fill = library_id) +
  geom_bar() +
  labs("All immune reference") +
  theme(legend.position = "bottom")

subset_reference_clusters <- has_cnv_df |>
  # bring in library_id
  dplyr::inner_join(
    infercnv_df |>
      dplyr::select(barcodes, library_id) |>
      unique()
  ) |>
  dplyr::filter(
    normal_reference == "subset-immune",
    subset_reference_cell
  )
subset_plot <- ggplot(subset_reference_clusters) +
  aes(x = infercnv_cluster, fill = library_id) +
  geom_bar() +
  labs("Subset immune reference") +
  theme(legend.position = "bottom")

wrap_plots(all_plot, subset_plot, guides = "collect") &
  theme(legend.position = "bottom")
```

How many CNVs were inferred per cluster?
We expect a single value here, because with `analysis_mode = "subclusters"` (the `inferCNV` default), CNV parameters are estimated per cluster.

@@

Now, we'll look at CNVs across chromosomes, including specific CNVs we expect to see in Ewing sarcoma:

@@

```{r}
cnv_validation_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    infercnv_cluster,
    total_cnv_per_cell,
    normal_reference,
    is_immune_reference,
    is_subset_reference,
    all_of(
      c(
        glue::glue("proportion_scaled_cnv_chr{1:22}"),
        glue::glue("proportion_scaled_loss_chr{1:22}"),
        glue::glue("proportion_scaled_dupli_chr{1:22}")
      )
    )
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, into = c("cnv_type", "chr")) |>
  dplyr::mutate(chr = stringr::str_remove(chr, "^chr"), chr = as.numeric(chr))
```
